---
title: "p8105_hw1_kpp2326"
author: "Kevin P. Patterson"
date: "2022-09-16"
output: html_document
---

```{r library_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```
## Problem 1

```{r load_data}
data("penguins", package = "palmerpenguins")
```

# Description
The penguins dataset contains `r nrow(penguins)` observations or individual penguin species. There are `r ncol(penguins)` variables, these variables detail the penguins' measurements, sex, year, species, and island. Upon closer examination, the categorical variables islands, years, and species contain 3 islands (Biscoe, Dream, and Torgersen), 3 years (2007, 2008, and 2009), and 3 species (Adelie, Chinstrap, and Gentoo), respectively. The mean flipper length of penguins with available flipper length data is `r round(mean(pull(penguins, flipper_length_mm), na.rm = T))` mm.

```{r scatterplot, echo=FALSE}
penguins %>%
  ggplot(aes(x=bill_length_mm,
             y=flipper_length_mm,
             color=species)) +
  geom_point() +
  xlab("Bill Length") + 
  ylab("Flipper Length") +
  labs(color= "Species", title = "Penguin Species Measurement Scatterplot")

ggsave("p8105_hw1_kpp2126_penguinscatterplot.pdf")
```
## Problem 2
```{r create_dataframe}
p2_df = 
  tibble(
    p2_sample = rnorm(10),
    p2_sample_gtr0 = p2_sample > 0,
    p2_sample_ch = c("a", "b", "c", "d", "e", "f", "g", "h", "i", "j"),
    p2_sample_fct = factor(c("low","low","low","med","med","med","high","high","high","high"))
  )
```

```{r mean}
p2_df %>%
  summarise(across(where(is.numeric), mean))

mean(pull(p2_df, p2_sample))
mean(pull(p2_df, p2_sample_gtr0))
mean(pull(p2_df, p2_sample_ch))
mean(pull(p2_df, p2_sample_fct))
```
I tried to take the mean for columns that contained numeric values and as expected it only produced a mean for `p2_sample`. 
In addition to tidyverse, I separately pulled the means for all columns and my results were successful for `p2_sample`, again, and for `p2_sample_gtr0`, which produces a result based on the values that are TRUE since I created a logical vector for all values greater than 0, resulting in a binary of 0 and 1.
For `p2_sample_ch` and `p2_sample_fct`, both resulted in warnings that indicated they're neither numeric nor logical, further stating an NA.

```{r convert_to_numeric, results='hide'}
p2_df %>%
  dplyr::mutate_at(c('p2_sample', 'p2_sample_gtr0', 'p2_sample_ch', 'p2_sample_fct'), as.numeric)

as.numeric(pull(p2_df, p2_sample))
as.numeric(pull(p2_df, p2_sample_gtr0))
as.numeric(pull(p2_df, p2_sample_ch))
as.numeric(pull(p2_df, p2_sample_fct))
```

I ran the conversion code first in tidyverse and separately using `as.numeric` for each variable, including the double coded variable `p2_sample` for consistency.
The `p2_sample_gtr0` converted the logical vectors to numeric values for 0 and 1, which can no longer be interpreted with the logical argument for false and true values and instead taken as literal 0 and 1 numeric values. 
The `p2_sample_ch` did not convert since they are characters and resulted in a warning and NAs.
The `p2_sample_fct` converted the previously coded factor levels to 1 (high), 2 (low) and 3 (med). The numbers were assigned based on alphabetical order.

This helps explains why the mean can't be taken when it's a character or a factor, because they are not coded as real numbers or logical vectors (denoting 0, false or 1, true).